---
title: Explore results of applying EBNMF to Montoro et al scRNA-seq data
author: Peter Carbonetto
output:
  workflowr::wflow_html:
    toc: no
    theme: readable
    highlight: textmate
    lib_dir: site_libs
    self_contained: no
---

Here I will create some plots to learn more about the EBNMF analysis
of the Montoro et al data set.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

Initial setup
-------------

Load some R packages that I will use to study the results:

```{r load-pkgs, message=FALSE}
library(ggplot2)
library(cowplot)
library(fastTopics)
```

Set the seed for reproducibility:

```{r set-seed}
set.seed(1)
```

Load the EBNMF results obtained from running the `montoro_fits.R`
script:

```{r load-results}
load("output/montoro_ebmf_fits.RData")
```

Extract some information about the cells:

```{r extract-meta-data}
x <- rownames(fl_k12$fl_ldf$F)
x <- strsplit(x,"_",fixed = TRUE)
meta_data <- data.frame(barcode   = sapply(x,"[[",1),
                        timepoint = sapply(x,"[[",2),
						mouse     = sapply(x,"[[",3),
						lineage   = sapply(x,"[[",4),
						celltype  = sapply(x,"[[",5))
meta_data <- transform(meta_data,
                       timepoint = factor(timepoint),
					   mouse     = factor(mouse),
					   lineage   = factor(lineage),
					   celltype  = factor(celltype))
```

EBNMF with 12 factors
---------------------

First is a plot of the first ("baseline") factor vs. the residual
s.d. parameters:

```{r residuals-k12, fig.height=2.15, fig.width=2.25}
residuals_plot <- function (baseline_exp, residual_sd) {
  pdat <- data.frame(baseline_exp = baseline_exp,
                     residual_sd = residual_sd)
  return(ggplot(pdat,aes(x = baseline_exp,y = residual_sd)) +
         geom_point(color = "black") +
		 labs(x = "baseline expression",y = "residual s.d.") + 
         theme_cowplot(font_size = 10))
}
residuals_plot(baseline_exp = fl_k12$fl_ldf$L[,1],
               residual_sd = fl_k12$residuals_sd)
```

These next plots summarize the priors on ${\bf F}$ and ${\bf L}$, and
could accompany the Structure plot shown below. However I'm not sure
if the scale parameter tells us anything useful—it is mainly the prior
probability that should inform the sparsity pattern. So maybe in the
end we will only use the left-hand plots.

The first two plots show the priors on the columns of ${\bf L}$.

```{r point-exp-priors-genes-k12, fig.height=1.75, fig.width=3.5}
factor_colors <- c(k1 = "black",k2 = "forestgreen",k3 = "royalblue",
                   k4 = "darkmagenta",k5 = "violet",k6 = "red",
				   k7 = "lightgray",k8 = "limegreen",k9 = "cyan",
				   k10 = "darkorange",k11 = "sienna",k12 = "gold")
fitted_priors_plot <- function (ghat, colors) {
  K <- length(ghat)
  pdat <- data.frame(k     = 1:K,
                     pi    = sapply(ghat,function (x) x$pi[2]),
                     scale = sapply(ghat,function (x) x$scale[2]),
					 color = colors)
  rows1 <- order(pdat$pi,decreasing = FALSE)
  rows2 <- order(pdat$scale,decreasing = FALSE)
  pdat1 <- pdat[rows1,]
  pdat2 <- pdat[rows2,]
  pdat1 <- transform(pdat1,k = factor(k,k))
  pdat2 <- transform(pdat2,k = factor(k,k))
  p1 <- ggplot(pdat1,aes(x = pi,y = k,color = k)) +
    geom_point() +
	scale_color_manual(values = pdat1$color) +
	labs(x = "prior probability") +
	guides(color = "none") +
    theme_cowplot(font_size = 9)
  p2 <- ggplot(pdat2,aes(x = scale,y = k,color = k)) +
    geom_point() +
	scale_color_manual(values = pdat2$color) + 
	labs(x = "scale parameter") +
	guides(color = "none") +
    theme_cowplot(font_size = 9)
  return(plot_grid(p1,p2,nrow = 1,ncol = 2))
 }
fitted_priors_plot(fl_k12$L_ghat,unname(factor_colors))
```

The second two plots show the priors on the columns of ${\bf F}$.

```{r point-exp-priors-cells-k12, fig.height=1.75, fig.width=3.5}
fitted_priors_plot(fl_k12$F_ghat,unname(factor_colors))
```

The non-negative matrix factorization with $K = 12$ factors identifies
the higher-level structures in the Montoro et al data set:

```{r structure-plots-k12, fig.height=5, fig.width=7.5, message=FALSE, results="hide"}
set.seed(1)
factors_to_plot <- c(7,4,5,11,8,2,6,9,10,3,12)
celltypes <- as.character(meta_data$celltype)
celltypes[celltypes == "Club (hillock-associated)"] <- "Club"
celltypes[celltypes == "Goblet.1"] <- "Goblet"
celltypes[celltypes == "Goblet.2"] <- "Goblet"
celltypes[celltypes == "Goblet.progenitor"] <- "Goblet"
celltypes[celltypes == "Tuft.1"] <- "Tuft"
celltypes[celltypes == "Tuft.2"] <- "Tuft"
celltypes[celltypes == "Tuft.progenitor"] <- "Tuft"
celltypes <- factor(celltypes,
                    c("Basal","Club","Goblet","Ionocyte","Ciliated",
					  "Tuft","Neuroendocrine","Proliferating"))
scale_cols <- function (A)
  return(t(t(A) / apply(A,2,max)))
L <- fl_k12$fl_ldf$L
L <- scale_cols(L)
F <- fl_k12$fl_ldf$F
F <- scale_cols(F)
colnames(L) <- paste0("k",1:12)
colnames(F) <- paste0("k",1:12)
rows <- c(sample(which(celltypes == "Basal"),10000),
          sample(which(celltypes == "Club"),5000),
          which(celltypes != "Basal" & celltypes != "Club"))
rows <- sort(rows)
F <- F[rows,]
celltypes <- celltypes[rows]
rows1 <- which(meta_data[rows,"timepoint"] == "Tp0")
rows2 <- which(meta_data[rows,"timepoint"] == "Tp30")
rows3 <- which(meta_data[rows,"timepoint"] == "Tp60")
p1 <- structure_plot(F[rows1,],grouping = celltypes[rows1],gap = 10,
                     topics = factors_to_plot,colors = factor_colors) +
  labs(y = "membership",title = "0 days") +
  guides(fill = "none") +
  theme(plot.title = element_text(size = 9,face = "plain"),
        plot.margin = unit(c(t = 0,r = 0,b = -2,l = 0),"lines"))
p2 <- structure_plot(F[rows2,],grouping = celltypes[rows2],gap = 10,
                     topics = factors_to_plot,colors = factor_colors) +
  labs(y = "membership",title = "30 days") +
  guides(fill = "none") +
  theme(plot.title = element_text(size = 9,face = "plain"),
        plot.margin = unit(c(t = 0,r = 0,b = -2,l = 0),"lines"))
p3 <- structure_plot(F[rows3,],grouping = celltypes[rows3],gap = 10,
                     topics = factors_to_plot,colors = factor_colors) +
  labs(y = "membership",fill = "factor",title = "60 days") +
  theme(plot.title = element_text(size = 9,face = "plain"),
        plot.margin = unit(c(t = 0,r = 0,b = 0,l = 0),"lines"))
plot_grid(p1,p2,p3,nrow = 3,ncol = 1,rel_heights = c(1,1,1.75))
```

EBNMF with 40 factors
---------------------

Plot first ("baseline") factor vs. the residual s.d. parameters:

```{r residuals-k40, fig.height=2.15, fig.width=2.25}
residuals_plot(baseline_exp = fl_k40$fl_ldf$L[,1],
               residual_sd = fl_k40$residuals_sd)
```

Plots summarizing the priors on the columns of ${\bf L}$:

```{r point-exp-priors-genes-k40, fig.height=4, fig.width=3.5}
factor_colors <-
  c(k1 = "black",k2 = "darkblue",k3 = "forestgreen",k4 = "blue",
    k5 = "cornflowerblue",k6 = "salmon",k7 = "gray",k8 = "black",
    k9 = "black",k10 = "limegreen",k11 = "black",k12 = "red",
    k13 = "darkorange",k14 = "black",k15 = "black",k16 = "dodgerblue",
    k17 = "royalblue",k18 = "black",k19 = "darkmagenta",k20 = "gold",
    k21 = "sienna",k22 = "black",k23 = "black",k24 = "black",
    k25 = "magenta",k26 = "black",k27 = "black",k28 = "black",
    k29 = "black",k30 = "black",k31 = "black",k32 = "black",
    k33 = "black",k34 = "black",k35 = "black",k36 = "black",
    k37 = "black",k38 = "black",k39 = "black",k40 = "black")
fitted_priors_plot(fl_k40$L_ghat,unname(factor_colors))
```

Plots summarizing the priors on the columns of ${\bf F}$:

```{r point-exp-priors-cells-k40, fig.height=4, fig.width=3.5}
fitted_priors_plot(fl_k40$F_ghat,unname(factor_colors))
```

Notice that the priors on ${\bf F}$ (which contains the memberships)
is *a priori* much more sparse than ${\bf L}$

The $K = 40$ fit captures more fine-scale structure. The Structure
plots show the primary cell-type-structure captured by the factors.
(Factors that capture other types of structure are not shown.) Here we
focus on the non-lineage-labeled cells ("tdTomato") since the
lineage-labeled cells feature fewer of the rare cell types such as
goblet and ionocyte.

```{r structure-plots-k40, fig.height=5.5, fig.width=7.5, message=FALSE, results="hide"}
set.seed(1)
celltype_factors <- c(7,19,2,3,4,5,21,17,16,13,6,12,10,25,20)
celltypes <- meta_data$celltype
celltypes <- as.character(meta_data$celltype)
celltypes[celltypes == "Club (hillock-associated)"] <- "Club (hillock)"
celltypes[celltypes == "Goblet.1"] <- "Goblet"
celltypes[celltypes == "Goblet.2"] <- "Goblet"
celltypes[celltypes == "Goblet.progenitor"] <- "Goblet"
celltypes[celltypes == "Tuft.1"] <- "Tuft"
celltypes[celltypes == "Tuft.2"] <- "Tuft"
celltypes[celltypes == "Tuft.progenitor"] <- "Tuft"
celltypes <- factor(celltypes,
                    c("Basal","Club","Club (hillock)",
                      "Goblet","Ionocyte","Ciliated",
                      "Tuft","Neuroendocrine","Proliferating"))
L <- fl_k40$fl_ldf$L
L <- scale_cols(L)
F <- fl_k40$fl_ldf$F
F <- scale_cols(F)
colnames(L) <- paste0("k",1:40)
colnames(F) <- paste0("k",1:40)
rows <- c(sample(which(celltypes == "Basal" &
                       meta_data$lineage == "Tom"),5000),
          sample(which(celltypes == "Club" &
                       meta_data$lineage == "Tom"),4000),
          which(celltypes != "Basal" & celltypes != "Club" &
                meta_data$lineage == "Tom"))
rows <- sort(rows)
F <- F[rows,]
celltypes <- celltypes[rows]
rows1 <- which(meta_data[rows,"timepoint"] == "Tp0")
rows2 <- which(meta_data[rows,"timepoint"] == "Tp30")
rows3 <- which(meta_data[rows,"timepoint"] == "Tp60")
p1 <- structure_plot(F[rows1,],grouping = celltypes[rows1],gap = 8,
                     topics = celltype_factors,colors = factor_colors) +
  labs(y = "membership",title = "0 days") +
  guides(fill = "none") +
  theme(plot.title = element_text(size = 9,face = "plain"),
        plot.margin = unit(c(t = 0,r = 0,b = -2,l = 0),"lines"))
p2 <- structure_plot(F[rows2,],grouping = celltypes[rows2],gap = 8,
                     topics = celltype_factors,colors = factor_colors) +
  labs(y = "membership",title = "30 days") +
  guides(fill = "none") +
  theme(plot.title = element_text(size = 9,face = "plain"),
        plot.margin = unit(c(t = 0,r = 0,b = -2,l = 0),"lines"))
p3 <- structure_plot(F[rows3,],grouping = celltypes[rows3],gap = 8,
                     topics = celltype_factors,colors = factor_colors) +
    labs(y = "membership",fill = "factor",title = "60 days") +
  ylim(0,1.75) +
  theme(plot.title = element_text(size = 9,face = "plain"),
        plot.margin = unit(c(t = 0,r = 0,b = 0,l = 0),"lines"))
plot_grid(p1,p2,p3,nrow = 3,ncol = 1,rel_heights = c(1,1,2))
```
