---
title: Yet more NMF analyses of the (noisy) swimmer data
author: Jason Willwerscheid and Peter Carbonetto
output:
  workflowr::wflow_html:
    toc: no
    theme: readable
    highlight: textmate
    lib_dir: site_libs
    self_contained: no
---

This workflowr page contains some additional explorations of NMF
methods—including the EBNMF methods implemented in flashier---for
learning parts from the "noisy" swimmer data.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

First, load the packages and some custom functions needed for the
analyses below.

```{r load-pkgs, message=FALSE}
library(R.matlab)
library(tibble)
library(dplyr)
library(tidyr)
library(ggplot2)
library(cowplot)
library(ebnm)
library(flashier)
source("code/swimmer_functions.R")
```

Initialize the sequence of pseudorandom numbers.

```{r set-seed}
set.seed(1)
```

Load the swimmer data set and convert the pixels to binary.

```{r load-swimmer}
Y <- readMat("data/swimmer.mat")$Y
Y <- apply(Y,3,as.vector)
Y <- Y - 1
Y[Y > 0] <- 1
```

Add some noise to the images.

```{r add-noise, fig.height=2, fig.width=7}
Y <- generate_noisy_swimmer_data(Y,0.05)
writeMat("noisy_swimmer.mat",Y = Y)
plot_images(Y[,1:16],nrow = 2,ncol = 8)
```

This is the decomposition produced by "vanilla NMF" with 16 factors—by
"vanilla", we mean NMF without any penalties on the parameters. See
[noisy_swimmer.m][noisy_swimmer_matlab] for details.

```{r vanilla-nmf, fig.height=2, fig.width=7}
nmf <- readMat("matlab/noisy_swimmer_nmf.mat")
plot_images(nmf$W,nrow = 2,ncol = 8)
```

This decomposition has trouble distinguishing the torso from the
limbs, and doesn't always split the limbs into separate factors. Also,
many of the limbs are quite noisy. Perhaps encouraging the parts to be
more sparse would address these issues with this. In this second
attempt (also implemented in [noisy_swimmer.m][noisy_swimmer_matlab]),
we forced the average sparsity of the 17 factors to be at least 0.8:

```{r sparse-nmf, fig.height=3, fig.width=7}
sparse_nmf <- readMat("matlab/noisy_swimmer_nmf_sW=0.8.mat")
plot_images(sparse_nmf$W,nrow = 3,ncol = 8)
```

Although not perfect—the factors have captured a lot of background
noise, and the torso is still included with some of the limbs—this
sparse NMF decomposition is nonetheless a clear improvement over the
vanilla NMF result.

Let's now see what happens when we force the sparsity to be even
higher (0.9):

```{r more-sparse-nmf, fig.height=3, fig.width=7}
more_sparse_nmf <- readMat("matlab/noisy_swimmer_nmf_sW=0.9.mat")
plot_images(more_sparse_nmf$W,nrow = 3,ncol = 8)
```

The sparser decomposition removes a lot of the background noise in the
factors, but also introduces some new problems: for example, some of
the factors are reduced to a single pixel or just a few pixels, and
other factors now include multiple limbs, which is not right.

To double-check, here are the average sparsity values from each of
these NMF decompositions:

```{r compare-sparsity}
mean(nmf$sp)
mean(sparse_nmf$sp)
mean(more_sparse_nmf$sp)
```

Let's now compare the NMF estimtes to a decomposition-by-parts
obtained by running flashier with sparse, non-negative priors
(`ebnm_point_exponential`). Remarkably, running flash with the
point-exponential prior gets the right result, automatically splitting
the torso and limb positions correctly into 17 factors, and adapting
the priors to capture the fact that the factors are very sparse (note
that we did not tell flashier that we wanted exactly 17 factors):

```{r flash, results="hide", fig.height=3, fig.width=7}
fit1 <- flash(Y,ebnm_fn = ebnm_point_exponential,greedy_Kmax = 18,
              backfit = TRUE,var_type = 0)
plot_images(ldf(fit1)$L,nrow = 3,ncol = 8)
```

ADD TEXT HERE.

[noisy_swimmer_matlab]: https://github.com/willwerscheid/ebnmf-paper/blob/master/matlab/noisy_swimmer.m
